The bhsdtr package: a general purpose method of bayesian inference for Signal Detection Theory models

### 2017-12-06 inst/preprint/luxsdt.tex
wersja ktora poszla do Mikea

% mixing properties, constrained czy restricted

\documentclass[oneside,a4paper]{article}
\usepackage[sectionbib,bibnewpage]{apacite}
\usepackage{amsmath}
\usepackage{bm}
\usepackage{float}
\usepackage{graphicx}
\usepackage{listings}

\title{The bhsdtr package: flexible bayesian inference for
  hierarchical or non-hierarchical equal variance normal Signal
  Detection Theory models with or without ratings.}

@@

We describe a novel method of bayesian inference for hierarchical or
non-hierarchical equal variance normal Signal Detection Theory
models. The method is implemented as an open-source R package that
uses state of the art platform Stan for sampling from the posterior
distribution. Our method can accommodate binary responses as well as
additional ratings, the user can specify an arbitrary number of random
grouping factors, sensitivity and criteria parameters can be regressed
on additional predictors within the same model via intermediate
unconstrained parameters, and the model can be extended by using an
automatically generated human-readable Stan code as a template. We
explain how our method improves other similar available methods, we
also provide an overview of the package and demonstrate the ease of
use and correctness of implementation by providing a real study data
analysis walk-through and by showing that the model successfully
recovers known parameter values when fitted to simulated data.

@@

Many tasks used in psychology studies are essentially classification
tasks. For example, in a memory study a test item may be either old or
new or in a perceptual study an object may either be a letter or a
digit. If a task requires classification it is always possible that
conclusions based on accuracy or percent correct are invalid because
the ability to discriminate between stimulus classes (i.e.,
sensitivity) is confounded with bias, which is the tendency to
classify stimuli as belonging to a particular class
\cite{GreenSwets66}. In principle, any effect that manifests itself in
the differences in classification accuracy may reflect differences in
sensitivity, bias or both. Signal Detection Theory provides a simple
and popular solution to this common and important problem. However,
SDT models are non-linear and variability in SDT parameters due to
factors such as participants or items has to be accounted for. When
effects of such factors are not accounted for in non-linear models
parameter estimates are biased and the associated uncertainty may be
underestimated.

We have created the \texttt{bhsdtr} package because in our opinion
available general purpose methods of fitting hierarchical SDT models
suffer from important limitations. The goal was to create a tool that
would be useful in a wide range of typical applications, that would
cover a broad spectrum of possible hierarchical structures defined on
SDT parameters and that would be relatively easy to use. We believe
that the \texttt{bhsdtr} package fulfills all of these goals. The
package is publicly available at
https://github.com/boryspaulewicz/bhsdtr, together with annotated R
code and data that was used to produce all the plots and perform all
the analyses described in this paper.

Our method is meant to be used by researchers already familiar with
Signal Detection Theory, hierarchical modelling and bayesian inference
so we begin by providing only a brief refreshment on certain issues in
SDT modelling. To those who do not yet possess the required expertise
we recommend \citeA{GelmanHill2007} for an introduction to bayesian
inference, hierarchical modelling and R and
\citeA{macmillan2004detection} for an introduction to SDT. Before we
go any further, however, a note on terminology seems in order.

In the context of hierarchical modelling factors such as participants,
items or replications are often referred to as groups, a naming
convention that we find confusing at times. For example, using this
terminology a single participant is both a group and a member of some
group, at the same time the term "group" is perhaps most strongly
associated with study conditions, as in "experimental group". In this
paper we use the term "sampled factor" instead, as it seems to capture
all the essential properties of such variables, i.e., the nominal
scale, the fact that values are sampled from a larger population and
are usually not of direct interest, as in "this is only a sample", and
that conclusions of statistical analysis are meant to apply to the
whole population of possible values (e.g., all the people or all the
items of a certain kind).

@@

A generalization of the basic SDT model that we focus on here is shown
in Fig.~\ref{fig:1} below.

\begin{figure}[H]
  \centering
  \includegraphics[width=.8\linewidth]{plot_sdtr.pdf}
  \caption{Equal variance normal Signal Detection Theory model with
    additional criteria}
  \label{fig:1}
\end{figure}

The stimulus is assumed to give rise, by some unspecified cognitive
process, to unidimensional internal evidence value $s$ sampled from a
distribution that depends on the stimulus class. For historical
reasons the stimulus classess are often referred to as "noise" and
"signal" and task performance is described in terms of hits, correct
rejections, omissions and false alarms, but this terminology is
appropriate only when the model is applied to tasks that require
detection, which is by far not always the case. In the most widely
used version of the model the two evidence distributions are normal
with the same variance which is usually fixed at unity to make the
model identifiable. Sensitivity is represented by the distance $d'$
between the means of evidence distributions. Because normal
distributions are unbounded evidence is always ambiguous and a
criterion $c$ placed on the evidence axis has to be used to reach a
discrete decision. The location of the decision criterion represents
the direction and magnitude of bias. The generalized version of the
model shown in Fig.~\ref{fig:1} is applicable in studies where
participants are asked to rate their binary classification decisions
on a performance or stimulus related dimension, confidence being a
prime example. The ratings and binary classification decisions can be
provided together (e.g., "I am almost certain that it was a digit"),
or in an arbitrary order (ratings before or after the decision).

The ratings are accommodated by modelling combined response $y$, which
represents both the binary classification decision and rating. The
value of $y$ increases with the strength of evidence in favor of the
second stimulus class. For example, if confidence is rated on a
four-point scale, then $y = 1$ when the subject decides that the
stimulus belongs to the first class (e.g., an old item) with certainty
4, $y = 2$ corresponds to first stimulus class with certainty 3,
$y = 5$ corresponds to second stimulus class (e.g., a new item) with
certainty 1, and $y = 8$ corresponds to second stimulus class with
certainty 4. More formally, the subject is assumed to give response
$k$ if $k = \min \{i : s <= c_i\}$, where
$s \sim \text{Normal}(\mu, 1)$ and $\mu = -d'/2$ for the first
stimulus class and $\mu = d'/2$ for the second stimulus class, $K$ is
the number of possible responses and $c_i$ are the decision criteria,
with $c_0$ and $c_K$ fixed at $-\infty$ and $+\infty$ respectively.

When $K = 2$ (no ratings) the model always fits perfectly, because the
data and the model have the same dimensionality. This makes
generalization to the $K > 2$ case particularly important, as it is
only when $K > 2$ that formal assumptions of the model (e.g., equal or
unequal variance) can be tested\footnote{In contrast to formal
  assumptions the \emph{psychological interpretation} of SDT
  parameters can be tested even when ratings are not available, e.g.,
  by means of selective influence \cite{sternberg2001separate}}, often
by comparing the observed and the theoretical ROC
curves. % , although to the best of our knowledge there exists no direct
% test of the ROC curve shape.  Off course, depending on research
% questions these additional criteria can be of independent interest,
% as can be the ratio of evidence distribution variances or their
% functional form, but here we concentrate on more basic applications.

SDT is a non-linear model, and when non-linear models are used in
psychology studies researchers are usually interested in the
relationships between the free parameters of such models and certain
additional measured or manipulated variables. A good example in the
context of classification performance is the dependence of $d'$ (a
free parameter of a non-linear SDT model) on stimulus strength
(usually a manipulated variable). Also, in great many psychology
studies data have hierarchical structure, e.g., there are repeated
measures for each participant or item and participants or items are
only samples from the target population. It seems worthwhile to
explain why it is important that our method accommodates both kinds of
situations.

\subsection{Importance of hierarchical structure}

If data have hierarchical structure, as is typically the case when SDT
models are used, and assuming that the SDT model is true, it is only
when the requirement of accounting for variability due to subjects,
items or other such factors is met that the estimates of the average
(fixed) effects are guaranteed to be unbiased and the conclusions are
guaranteed to generalize to the target population.

A not so uncommon practice of analysing data aggregated over subjects,
items or other sampled factors represents an extreme case of ignoring
hierarchical data structure. Invalidity of this approach in the
context of SDT was clearly illustrated with results of simulational
studies by \citeA{morey2008problematic}, although strictly speaking it
requires such demonstrations only in order to make a point about a
specific case. Because SDT is a non-linear model, by definition
expected values of estimates of its parameters based on data
aggregated over sampled factors are not in general true population
averages. Such estimates are biased and inference about the target
population based on such estimates is simply not valid. To give a
concrete example, concider two unbiased participants, one with
$d' = 2$ and one with $d' = 4$. Their expected average accuracy is
given by $(\Phi(1) + \Phi(2)) / 2 = .91$, which corresponds to
$d' = 2.68$, whereas their true average $d'$ is $3$.

Aggregation is not the only way of ignoring hierarchical data
structure. Sometimes non-aggregated data are analyzed by using
separate estimates for every participant $\times$ item $\times$
condition combination but uncertainty due to subject or item effects
\emph{distributions} is not accounted for by means of hierarchical
model structure. In this case the conclusions, at least with respect
to the uncertainty in the estimates, are guaranteed to be valid only
for the given sample, not the target population. This is rarely, if
ever, the purpose of a study.

\subsection{Importance of including regression structure}

Another common source of problems is separation of estimation of
non-linear model parameters from regression analysis relating these
parameters to other measured or manipulated variables. When SDT
parameters are estimated separately for each subject and condition and
only later regressed on predictors of interest a number of additional
issues may arise.

Firstly, standard errors or credible intervals associated with
regression coefficients do not reflect the uncertainty in SDT
parameter estimates, because the latter are treated as mere data
points. This may lead to overly optimistic inference on regression
coefficients, since what this amounts to, among other things, is
ignoring part of the residual variance.

Secondly, including the regression part in the same model makes the
SDT parameter estimates dependent on the common regression structure
and so also on each other. This can improve their quality and affect
estimates of their uncertainty, just like assuming that random effects
are samples from a common distribution may improve their estimates by
making them dependent in the model.

Thirdly, the precision of parameter estimates often varies between
participants, items or conditions but when these estimates are treated
as mere data points in regression analysis no use is made of this
information. When regression structure is included in the same model
the regression coefficients are less influenced by the parameter
estimates that have lower precision.

\section{A critical summary of avaiable methods of bayesian inference
  for hierarchical SDT models}

To the best of our knowledge only two attempts at creating something
akin to the general purpose method of inference for hierarchical SDT
models with ratings can be found in literature. We will not consider
the ROC Toolbox \cite{Koen2016} or other similar tools here because as
of this writing these tools cannot be used to fit hierarchical SDT
models. The two available methods that can be used to fit hierarchical
SDT models are the Gibbs sampler proposed by
\citeA{morey2008problematic} and the Hierarchical Meta-d' model
(HMeta-d') proposed by \citeA{hmetad}. In our summary of these two
methods we mainly concentrate on two issues, namely the correctness of
hierarchical structure implementation and possibility to model SDT
parameters as dependent on additional variables.

\citeA{morey2008problematic} proposed a Gibbs sampling algorithm for
inference on hierarchical unequal variance normal SDT models with
ratings. A slightly modified version of the algorithm was later made
available as part of the \texttt{hbmem} R package. The authors adopted
what they called a fixed-criteria parametrization. Under this
parametrization the outermost criteria $\bm{c}_1$ and $\bm{c}_{K-1}$
are fixed at $0$ and $1$ respectively, while the remaining criteria
together with means and variances of the evidence distributions are
free to vary. The hierarchical part of evidence distribution means is
represented by additive independent normal effects of at most two
sampled factors which, following the authors, we label as subject and
item effects for convenience. However, the criteria are represented as
subject specific \emph{fixed}, not random effects. Items can have an
indirect and somewhat limited random effect on the criteria, because
they can have an effect on evidence distribution means by shifting
them independently. As the authors clarify, shifting both means by the
same amount in the same direction does not change the $d'$ but
simultaneously shifts all the criteria relative to the evidence
distributions. This allows an item to have the same effect on all the
criteria but it does not allow an item to have different effects on
different criteria. Random effects correlations are not accounted
for. Criteria are order-restricted, but the means of evidence
distributions are order-restricted only in the version implemented in
the \texttt{hbmem} package. SDT parameters can be regressed on
additional predictors to some extent by specifying a model
matrix. Lastly, because this is a Gibbs sampler, authors rely on
conjugate priors, including the problematic \cite{gelman2004prior}
inverse gamma prior placed on variances.

Fleming's \citeyear{hmetad} HMeta-d' model is a hierarchical version
of the meta-d' model \cite{maniscalco2012signal}, which in turn is a
generalization of the SDT model with additional criteria that allows
for a separate "meta-sensitivity" to account for possible
discrepancies between binary stimulus classification (referred to as
type 1 task) and a rating task (referred to as type 2 or
meta-cognitive task). We consider HMeta-d' here because it reduces to
the SDT model with ratings when type 1 and type 2 sensitivities are
equal. In HMeta-d' evidence distributions are normal with equal
variance. Normally distributed random effects of one factor are
allowed for sensitivities and parameters indirectly representing the
criteria, for lack of a better name from now on referred to as raw
criteria. The actual criteria are subject-specific sorted values of
unconstrained raw criteria and it is these unconstrained parameters,
not the actual criteria, that are given independent hierarchical
normal distributions representing variability due to subjects. Since
sorting is not injective the space of raw criteria is not isomorphic
to the space of actual criteria and each element of the raw criteria
vector can represent a value of any criterion, depending on the values
of other elements of the raw criteria vector. The model does not
account for possible correlations between random
effects. Sensitivities are represented as unrestricted scalars even
though they are non-negative by definition. Out of the box the code
provided by the author allows only for a very limited regression
structure to be placed on SDT parameters. Software implementation
relies on ready-made model specifications expressed in JAGS language
that are aimed at several simple types of experimental design (e.g.,
single subject or single group of subjects). The models can be fitted
using convenient MATLAB functions.

\section{Hierarchical Signal Detection Theory in unconstrained
  parameter space}

As we have already stressed, in a typical study both sensitivity and
criteria parameters have to be supplemented with hierarchical
structure for inference to be guaranteed to generalize to the target
population. In our critical summary of the two available methods we
have listed some reasons why we believe that without substantial
modifications neither method correctly addresses this issue. The main
reason why both methods fail in correctly specifying the hierarchical
structure on SDT parameters is that random effects are represented by
normal distributions but parameters are constrained, i.e.,
sensitivities are non-negative and criteria are ordered.

In the \texttt{bhsdtr} package sensitivities and criteria are derived
from unconstrained parameters by convenient isomorphisms. This is
easily achieved in case of sensitivity by deriving $d'$ from
$\delta = \ln(d')$ and modelling random effects on $d'$, if any, by
assuming that $\delta$ is normally distributed. Difficulties in
representing the hierarchical structure of the criteria stem from the
fact that even though each criterion considered in isolation is an
unconstrained parameter, the vector of all the criteria is ordered.
We propose to map the $R^{K-1}$ space, representing unconstrained
criteria, to the $K$ dimensional probability simplex space using the
softmax function and to map the simplex space to the ordered criteria
space by means of the inverse normal CDF. Specifically, assume that a
normal distribution is centered on the midpoint between the two
evidence distributions. The value of criterion $i$ is then given by:

@@

\noindent where $\bm{\gamma} \in R^K$ with $\bm{\gamma}_K$ fixed at
$0$ for identifiability. The idea is illustrated in
Fig.~\ref{fig:2} below:

\begin{figure}[H]
  \centering
  \includegraphics[width=.8\linewidth]{plot.pdf}
  \caption{Mapping between unconstrained $\gamma$ vector and criteria}
  \label{fig:2}
\end{figure}

>>>>

% WĄTPLIWOŚCI JĘZYKOWE oznaczone przez %?? !! oznacza miejsce, gdzie
% skończyłem ostatnio poprawiać

% przykłady w sieci value of <> [directly] represents the <>

% jak się pisze 16ms czy 16 ms

% argumenty funkcji są an, gdy piszemy ogólnie o działaniu tych
% funkcji

\documentclass[oneside,a4paper]{article}
\usepackage[sectionbib,bibnewpage]{apacite}
\usepackage{amsmath}
\usepackage{bm}
\usepackage{float}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{caption}

\title{The bhsdtr package: a general purpose method of bayesian
  inference for Signal Detection Theory models}

@@

We describe a novel method of bayesian inference for hierarchical or
non-hierarchical equal variance normal Signal Detection Theory models
with one or more criteria. The method is implemented as an open-source
R package that uses the state-of-the-art platform Stan for sampling
from posterior distributions. Our method can accommodate binary
responses as well as additional ratings and an arbitrary number of
nested
% ?? WSZYSTKIE sdt parameters CAN be regressed, więc the
or crossed random grouping factors. The SDT parameters can be
regressed on additional predictors within the same model via
intermediate unconstrained parameters, and the model can be extended
by using automatically generated human-readable Stan code as a
% ?? czy to jest przypadek other -> bez przedimka?
template. In the paper we explain how our method improves on other
similar available methods, we give an overview of the package,
demonstrate the ease of use by providing a real-study data analysis
walk-through, and show that the model successfully recovers known
parameter values when fitted to simulated data.

@@

Many tasks used in psychology studies are essentially classification
tasks. For example, in a memory study participants may be required to
decide if a given test item is old or new, or, in a perceptual study,
an object may either be a letter or a digit. If a task requires
classification it is always possible that conclusions based on
% ? an ability mi tu nie pasuje
accuracy or percent correct are invalid because ability to
discriminate between stimulus classes (i.e., sensitivity) is
confounded with bias, which is a tendency to classify stimuli as
belonging to a particular class \cite{GreenSwets66}. In principle, any
effect that manifests itself in differences in classification accuracy
% ?? ani the sensitivity ani a sensitivity nie pasuje. Chodzi o
% ogólnie rozumiane zjawisko.
may reflect differences in sensitivity, bias, or both. Signal
Detection Theory provides a simple and popular solution to this common
and important problem.

However, because the SDT model is non-linear any variability in its
parameters due to factors such as participants or items has to be
accounted for, otherwise the estimates of SDT parameters are
biased. As we explain later in this paper, neither of the two
available methods of inference for hierarchical SDT models that we are
aware of addresses this problem correctly. Hence, our main goal was to
create a correct implementation of the general hierarchical linear
regression structure defined on SDT parameters. We argue that the
\texttt{bhsdtr} package for R \cite{rstatistical} provides exactly
such an implementation and we have made it publicly available at
% ?? tylko jeden, konkretny kod i zbiór danych, o którym czytelnik
% dopiero się dowiaduje, ale o którym nie będziemy więcej mówić, czyli
% the
https://github.com/boryspaulewicz/bhsdtr, together with the annotated
source code that was used to perform all the analyses and produce all
the figures presented in this paper.

% OPIS STRUKTURY ARTYKUŁU

In what follows, after introducing the most common version of the SDT
model, we describe its generalization which can accomodate data from
rating experiments. Next, we explain briefly why, if a method of
inference for SDT models were to be of general use in psychology
studies, it is essential that it was based on a model equipped with
the general hierarchical linear regression structure. The
\texttt{bhsdtr} package meets this requirement thanks to a novel
parametrization. We describe this novel parametrization in great
detail and we motivate it by explaining how the reliance on a
relatively standard parametrisation leads to problems in the two other
available implementations. We end the first part of this paper with a
formal definition of the model as implemented in \texttt{bhsdtr}. The
second part contains an overview of the package and a tutorial in
which we demonstrate how to use our method in practice. Before we go
any further, however, a note on terminology seems in order.

% SAMPLED FACTOR

In the context of hierarchical modelling factors such as participants,
items, or replications are often referred to as groups. In our opinion
this naming convention may be confusing; a single participant is both
a group and a member of some group, while at the same time the term
"group" is perhaps most strongly associated with study conditions, as
in "experimental group". In this paper we use the term "sampled
factor" instead, because, by virtue of being a new term, it is
unambiguous, and because it seems descriprively correct: the term
"sampled factor" captures all the essential properties of such
variables, i.e., a nominal scale, the fact that values are sampled
from a larger population and are usually not of direct interest, as in
"this is only a sample", and that conclusions
% ?? jeżeli już mamy jakąś analizę statystyczną, to zakładamy, w
% uproszczeniu, że target population jest określona
of statistical analysis are meant to apply to the whole population of
possible values.

@@

According to Signal Detection Theory each stimulus in a classification
task gives rise, by some unspecified cognitive process, to an
unidimensional internal evidence value $s$ sampled from
% ?? stimulus już był wprowadzony, więc jego klasa jest (jakaś, ale)
% określona
a distribution that depends on the stimulus class. For historical
reasons the two stimulus classess are often referred to as "noise" and
"signal", and task performance is described in terms of hits, correct
rejections, omissions, and false alarms, but this terminology is
appropriate only when the model is applied to tasks that require
detection, which is by far not always the case. In the most widely
used version of the model, shown in Fig.~\ref{basicsdt}, the two
evidence distributions are normal with the same variance, which is
usually fixed at unity to make the model identifiable. The distance
$d'$ between the means of the
% ?? sensitivity bez przedimka?
evidence distributions represents sensitivity. Because normal
distributions are unbounded, $s$ is always ambiguous, and so a
criterion $c$ placed on the evidence axis has to be used to reach a
discrete
% ?? bias bez przedimka?
decision. A participant is assumed to decide that a stimulus belongs
to the first class (e.g., an old item) if $s < c$, or that it belongs
to the second class (e.g., a new item) if $s \geq c$. The location of
the decision criterion $c$ represents bias.

\begin{figure}[H]
  \centering
  \includegraphics[width=.8\linewidth]{plot_basicsdt.pdf}
  \caption{Equal variance normal Signal Detection Theory model}
  \label{basicsdt}
\end{figure}

% Najprostsze użycie

Perhaps the simplest way of using this model is to fit it to observed
response counts and use the estimated $d'$ values in place of the
percent correct scores; If the model is correct, the resulting
performance measure is not contaminated by bias. Obviously, the model
may not be correct, which is one of the reasons why we focus more on
the generalized version shown in Fig.~\ref{sdtr} below.

% Model K >= 2

\begin{figure}[H]
  \centering
  \includegraphics[width=.8\linewidth]{plot_sdtr.pdf}
  \caption{Equal variance normal Signal Detection Theory model with
    additional criteria}
  \label{sdtr}
\end{figure}

This
% ?? applicable to czy applicable in
generalized model is applicable to studies where participants are
asked to rate their binary classification decisions on confidence or
some other performance or stimulus related
% the ratings, bo study już określone, podobnie jak (rodzaj)
% klasyfikacji binarnej, ale czy the ratings and THE binary ...?
dimension. The ratings and the binary classification decisions can be
provided together (e.g., "I am almost certain that it was a digit"),
or in an arbitrary order.

% MODELOWANIE RATINGÓW ZA POMOCĄ KRYTERIÓW

% ?? a combined response? Skoro mamy ratings (i w domyśle mamy też
% binary class.), to czy nie the combined response? A może
% jakiekolwiek ratingi są załatwione za pomocą (ogólnego rozwiązania)
% modelowania czegoś, co jest "odpowiedzią łączną"?
Ratings are accommodated by introducing additional criteria and
modelling a combined response $y$, which
% ?? znowu kwestia przedimków vs. and
represents both the binary classification decision and rating. The
value of $y$ increases with the strength of evidence in favor of the
second stimulus class. For example, if confidence is rated on a
four-point scale, then $y = 1$ when a subject decides
% ?? bodziec, który zobaczył badany, więc the? bo chodzi o decyzję o
% określonym dla decydenta bodźcu, a nie nieokreślonym dla decydenta?
that a stimulus belongs to the first class with certainty $4$, and
$y = 5$ when a subject decides that a stimulus belongs to the second
class with certainty $1$. More formally, if $K$ is the number of
possible combined responses then a subject is assumed to give response
$k$ if $s \in (\bm{c}_{k-1},\bm{c}_k]$, where $\bm{c}$ are the $K+1$
decision criteria, with $\bm{c}_0$ and $\bm{c}_K$ fixed at $-\infty$
and $+\infty$ respectively.

% UŻYWAĆ SDTR CHOĆBY PO TO, ŻEBY SPRAWDZIĆ JEGO ZAŁOŻENIA

There is a good reason to collect the ratings and use the generalized
SDT model from Fig.~\ref{sdtr} even when neither the ratings nor the
placement of criteria are relevant to the research problem.
% ?? If K = 2 to jakby: w (jakimś) przypadku, w którym, When K = 2 to
% jakby: zawsze, gdy ...? Czuję, że lepiej the data i the model potem
When $K = 2$ (no ratings), the SDT model fits perfectly, regardless of
it being a reasonably good approximation to reality or not, because
the data and the model have the same dimensionality. This makes the
generalization to the $K > 2$ case particularly important, as it is
% ?? niektóre lub wszystkie formal assumptions, ale mamy na myśli
% konkretny zbiór formal assumptions, który czytelnik już zna, a
% nawet, jeśli nie, to te założenia są unikalnie wyznaczone przed
% model - the?
only when $K > 2$, that the formal assumptions of the model (e.g.,
equal or unequal variance) can be tested\footnote{In contrast to the
  formal assumptions, a psychological interpretation of the SDT
  parameters can be tested even when ratings are not available, e.g.,
  by means of selective influence \cite{sternberg2001separate}}, which
is often done by comparing observed and predicted ROC curves.

When SDT models are used in psychology studies, researchers are
usually interested in relationships between SDT parameters and
additional measured or manipulated
% ?? the dependence, bo ogólnie rozumiana kwestia zależności, stimulus
% strength - ogólne zjawisko, a nie jakaś konkretna, albo pojedyncza,
% ale bliżej nieokreślona
variables; A good example is the dependence of $d'$ on stimulus
strength. Also, in a great majority of psychology studies in which
classification tasks are used the data have hierarchical structure,
i.e., there are repeated measures for each participant or item and
participants or items are only samples from the target population. It
seems worthwhile to explain why a general-purpose method of inference
for SDT models should accomodate both kinds of situations.

% ?? Ważność hierarchical structure jako takiej
\subsection{The importance of hierarchical regression structure}

If data have hierarchical structure, but variability due to subjects,
items, or other sampled factors is not accounted for, estimates of
average (fixed) effects are not guaranteed to be unbiased and
conclusions are not guaranteed to generalize to a target population. A
not so uncommon practice of analysing data aggregated over sampled
factors represents an extreme case of ignoring
% ?? nieokreślona hier. structure, bo tu mówimy o externe case
% ogólnego zjawiska ignorowania ogólnej kwestii hierarchiczności
hierarchical data structure. Invalidity of this approach in the
% ?? Jeżeli illustrated w czasie przeszłym dokonanym i mówimy zaraz
% czyje rezultaty, to chyba the results?
context of SDT was clearly illustrated with the results of
simulational studies by \citeA{morey2008problematic}, although,
strictly speaking, it requires such demonstrations only in order to
make a point about a specific case. Because SDT is a non-linear model,
% ?? data jest unikalne dla tokenów estimates, więc the data?
by definition, when estimates of its parameters are based on data
aggregated over sampled factors the expected values of these estimates
are not in general true population averages. Such estimates are
% ?? zakładamy (w uproszczeniu), że target population jest ustalona
% dla każdego estimate
biased and inference about a target population based on such estimates
is simply not valid. To give a concrete example, consider two unbiased
participants, one with $d'_1 = 2$ and one with $d'_2 = 4$. Their
expected average accuracy is given by
$(\Phi(d'_1/2) + \Phi(d'_2/2)) / 2 = .91$, which corresponds to
$d' = 2.68$, whereas their true average $d'$ is $3$.

Aggregation is not the only way of ignoring hierarchical data
structure. Sometimes non-aggregated data are analyzed by using
separate estimates for every participant $\times$ item $\times$
% ?? mamy już dane i estimates, to (zwykle nieznany) (poziom)
% uncertainty jest określony
condition combination, but the uncertainty due to subject or item
effects \emph{distributions} is not accounted for by means of a
hierarchical model
% dalej jedziemy z nieokreślonymi, bo wyjaśniamy ogólne zasady:
% wnioski mogą być jakiekolwiek, próbki i populacje są bliżej
% nieokreślone
structure. In such cases conclusions, at least with respect to
% ?? może chodzić o różne population-average różnie estymowalne
%
% ?? population-average - or fixed - effects? chodzi mi o to, że
% effects ma się też "łączyć" z population-average
the uncertainties in estimates of population-average (fixed)
% ?? sample i population jest dana dla tokenów
effects, are guaranteed to be valid only for the given sample, not the
target population.

% mówimy o ogólnym co prawda, ale zarazem pojedynczym/unikalnym
% problemie: the separation? Chętnie bym się pozbył przedimka
Another common source of problems is the separation of estimation of
non-linear model parameters from regression analysis whichs aim to
relate these parameters to measured or manipulated variables. When the
SDT parameters are estimated separately for each subject and
condition, and only later regressed on predictors of interest, a
number of additional issues may arise.

% tutaj jakaś regression analysis już jest wcześniej wprowadzona
Firstly, the standard errors or credible intervals associated with the
regression coefficients do not reflect the uncertainty in the SDT
parameter estimates, because the latter are treated as mere data
points. Precision of parameter estimates often varies between
participants, items, or conditions, but when the estimates are treated
as data points no use is made of this information. Secondly,
regressing parameters on numeric predictors makes their estimates
dependent on the common regression structure, and so also on each
other, which can improve the quality of the estimates, just as
assuming that random effects are samples from a common distribution
may improve their estimates.

The aforementioned problems can be dealt with by supplementing the SDT
model with the general hierarchical linear regression structure. For
the reasons that we will now explain, this can only be achieved if the
model is substantially reparametrised.

% dopiero wprowadzamy ową unconstrained parameter space
\section{Hierarchical Signal Detection Theory in a constrained
  parameter space}

Both $d'$ and $c$ have the virtue of being directly interpretable in
terms of sensitivity and bias. However, both $d'$ and $c$ are
constrained: $d'$ is non-negative and, when there is more than one
criterion, the elements of the $\bm{c}$ vector are order-restricted
($\bm{c}_{i+1} > \bm{c}_i$). Because normal distribution is unbounded,
hierarchical linear regression structure, in which random effects are
normally distributed and fixed effects are represented by
unconstrained parameters, can only be defined on unconstrained
parameters. We provide examples of the problems that may otherwise
arise in the following summary of hierarchical SDT implementations.

We are aware of only two attempts at creating a general purpose method
of inference for hierarchical SDT models with ratings. One is the
Gibbs sampler proposed by \citeA{morey2008problematic} and the other
is the Hierarchical Meta-d' model (HMeta-d') proposed by
\citeA{hmetad}.
% We do not consider the ROC Toolbox \cite{Koen2016} or other similar
% tools here because, as of this writing, these tools cannot be used to
% fit hierarchical SDT models. The two available methods that can be
% used to fit hierarchical SDT models are
The HMeta-d' model is a hierarchical version of the meta-d' model
\cite{maniscalco2012signal}, which in turn is a generalization of the
SDT model that allows for a separate "meta-sensitivity" to account for
possible discrepancies
% referred to as ... bez przyimka, prawda?
between a binary stimulus classification (referred to as type 1 task)
% ?? wcześniej było a binary i a rating, jakby to były (potencjalnie)
% niezwiązane zadania
and the associated rating task (referred to as type 2 or
meta-cognitive task). We consider HMeta-d' here because it reduces to
the SDT model with
% ?? powtórzenie the
ratings when the type 1 and the type 2 sensitivities are equal.

The Gibbs sampler created by \citeA{morey2008problematic} allows for
at most two sampled factors to have independent normally distributed
random effects on the evidence distribution means. Unlike $d'$, each
evidence distribution mean considered in isolation is an unconstrained
parameter, but the mean of the second one is by definition greater
($d' > 0$), or equal ($d' = 0$), than the mean of the first one. The
authors explicitly admit that their algorithm does not enforce this
restriction, because, as they claim in the paper, it would greatly
complicate analysis. At the same time the authors fail to mention that
an immediate consequence of allowing for negative $d'$ values is that
the resulting posterior distribution looses its intended
interpretation, since it can have a non-zero mass when $d' < 0$. The
outermost criteria are fixed at $0$ and $1$, and the oredering
restriction is enforced by assuming that the likelihood is $0$
whenever $\bm{c}_{i+1} \leq \bm{c}_i$. As the authors explain, because
a sampled factor can have independent random effects on the evidence
distribution means, it can have an effect on all the criteria:
shifting both means by the same amount in the same direction is
equivalent to keeping the sensitivity intact, while shifting the
criteria relative to the evidence distributions. However, the elements
of the criteria vector cannot be affected differently by the same
sampled factor, which is an unrealistic restriction.

In HMeta-d' the hierarchical structure is restricted to normally
distributed random intercepts of one sampled factor. Perhaps the most
problematic aspect of the HMeta-d' model, apart from the fact that
also in this model the $d'$ parameter is allowed to assume negative
values, is the representation of the criteria. Each element of the
criteria vector has an associated independent normal distribution,
which does allow for criteria random effects, but does not enforce the
necessary ordering restriction. The elements of this vector are sorted
to obtain another criteria vector, and it is this sorted criteria
vector that is used to model the conditional combined response
distributions. Consequently, the model does contain parameters
representing the actual, order-restricted criteria, but, because
sorting is not injective, the
% ?? (i.e., not isomorphic to)
space of the actual criteria is only loosely related (i.e., not
isomorphic) to the space of the unrestricted criteria vectors that are
associated with a hierarchical structure.

% Hierarchical structure does not have to be linear, but when it is, the
% parameters associated with unconstrained fixed effects and normally
% distributed random effects of sampled factors cannot be non-negative
% or order-restricted. We propose to substantially reparametrise the SDT
% model.

\section{Hierarchical Signal Detection Theory in an unconstrained
  parameter space}

The general hierarchical linear regression structure can be defined on
SDT parameters only if the latter are derived from unconstrained
parameters. In the \texttt{bhsdtr} package $d'$ is derived from
$\delta = \ln(d')$, this way random effects on $d'$ can be modelled by
assuming that $\delta$ is normally distributed. The problem of
representing the criteria by unconstrained parameters
% ?? skoro the space of, to potem the... ? w końcu chodzi nam o
% *wszystkie*, takie, że...?
is solved by mapping the $R^{K-1}$ space of unconstrained criteria
vectors to the $K$ dimensional probability simplex space using the
softmax function, and mapping the simplex space to the space of
order-restricted criteria vectors by means of the inverse normal CDF:

@@

\noindent where $\Phi$ is the CDF of the standard normal distrubution
and $\bm{\gamma} \in R^K$, with $\bm{\gamma}_K$ fixed at $0$ for
identifiability. The idea is illustrated in Fig.~\ref{fig:2} below:

\begin{figure}[H]
  \centering
  \includegraphics[width=.8\linewidth]{plot.pdf}
  \caption{Mapping between the unconstrained $\bm{\gamma}$ vector and
    the criteria}
  \label{fig:2}
\end{figure}
